RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-06-27 background-full.tex
many improvements (see email)

Most of the examples in this article, as well as the topic of discussion in the transcripts we analyze, come from the traffic simulator design exercise. In this exercise, designers are provided a problem description, requirements, and a description of the desired outcomes. The problem description is given in full in Appendix~\ref{ch:designprompt}, %TODO SG: shown as ??
 and is summarized as follows: The client of the project is Professor E, who teaches civil engineering at UCI. %TODO SG: Make it anonymized. For example use a university in California (if this is UC Irvine)
 It is the task of the designer to specify a system in which the professor can teach students how various theories around traffic lights works, such as queuing theory. To this end, a piece of software has to be developed in which students can create visual maps of an area, regulate traffic, and so forth. The original version of the problem descrption~\cite{UCIworkshop} is well known in the field of design reasoning since it has been used in a workshop\footnote{\url{http://www.ics.uci.edu/design-workshop/}}, and transcripts of this workshop have been analyzed in detail~\cite{Petre:2013:SDA:2535028}. Although the concepts of traffic lights, lanes, and intersections are common and appear to be simple, building a traffic simulator to represent these relationships and events in real time turns out to be challenging.

@@

@@

Figure~\ref{fig:trafficsim} illustrates a GRL diagram from the traffic simulator design exercise. An actor (\includegraphics[scale=1]{img/actor}) represents a stakeholder of a system (\texttt{Student}, Figure~\ref{fig:trafficsim}), or the system itself ( \texttt{Traffic Tycoon}, Figure~\ref{fig:trafficsim}). Actors are holders of intentions; they are the active entities in the system or its environment who want goals to be achieved, tasks to be performed, resources to be available, and softgoals to be satisfied. Softgoals (\includegraphics[scale=1]{img/softgoal}) differentiate themselves from goals (\includegraphics[scale=1]{img/goal}) in that there is no clear, objective measure of satisfaction for a softgoal whereas a goal is quantifiable, often in a binary way. Softgoals (e.g.  \texttt{Realistic simulation)} are often related to non-functional requirements, whereas goals (such as  \texttt{Generate cars}) are related to functional requirements. Tasks (\includegraphics[scale=1]{img/task}) represent solutions to (or operationalizations of) goals and softgoals. In Figure~\ref{fig:trafficsim}, some of the tasks are  \texttt{Create new cars} and \texttt{Keep same cars}. In order to be achieved or completed, softgoals, goals, and tasks may require resources (\includegraphics[scale=1]{img/resource}) to be available (e.g., \texttt{External Library}, Figure~\ref{fig:trafficsim}).

\begin{figure*}[ht]
\centering
\includegraphics[width=\textwidth]{img/transcript_grl_0}
\caption{GRL Model for the traffic simulator.}
\label{fig:trafficsim}
\end{figure*}

Different links connect the elements in a GRL model. AND, IOR, and XOR decomposition links (\includegraphics[scale=1]{img/decomposition}) allow an element to be decomposed into sub-elements. In Figure~\ref{fig:trafficsim}, the goal  \texttt{Generate cars} is XOR-decomposed to the tasks \texttt{Create new cars} and \texttt{Keep same cars}. Contribution links (\includegraphics[scale=1]{img/contribution}) indicate desired impacts of one element on another element. A contribution link has a qualitative contribution type or a quantitative contribution. Task  \texttt{Create new cars} has a \emph{help} qualitative contribution to the softgoal  \texttt{Dynamic simulation}. Dependency links (\includegraphics[scale=1]{img/dependency}) model relationships between actors. For example, actor  \texttt{Traffic Tycoon} depends on the actor  \texttt{Student} to perform the task  \texttt{Adjust car spawning rate} to fulfill its task  \texttt{Generate cars}.

GRL is based on $i*$~\cite{Yu:1997:TMR:827255.827807} and the NFR Framework~\cite{chung2012non}, but it is not as restrictive as $i*$. Intentional elements and links can be more freely combined, the notion of agents is replaced with the more general notion of actors, i.e., stakeholders, and a task does not necessarily have to be an activity performed by an actor, but may also describe properties of a solution. GRL has a well-defined syntax and semantics, which are necessary if we want to incorporate it into a formal framework (requirements 1 and 2 as described in the introduction). Furthermore, GRL provides support for providing a scalable and consistent representation of multiple views/diagrams of the same goal model (see~\cite[Ch.2]{Ghanavati2013} for more details). GRL is also linked to Use Case Maps via URNLink ((\includegraphics[scale=1]{img/urnlink}) which provides traceability between concepts and instances of the goal model and behavioral design models. Multiple views and traceability are a good fit with our current research: we aim to add traceability links between intentional elements and their underlying arguments.

@@

The GRL model in Figure~\ref{fig:trafficsim} shows the softgoals, goals, tasks and the relationship between the different intentional elements in the model. However, the rationales and arguments behind certain intentional elements are not shown in the GRL model. Some of the questions that might be interesting to know about are the following:

\begin{itemize}
	\item Why does actor \texttt{Teacher} have only a single softgoal \texttt{Students learn from practice}? Why is this, for instance, not connected to any of the elements of \texttt{Student}?
	\item What does  \texttt{Adjust timing schemes of sensorless interactions} mean?
	\item Why does task \texttt{Keep same cars} contribut positively to \texttt{Realistic simulation} and negatively to \texttt{Dynamic simulation}?
	\item How does the \texttt{Student} control the \texttt{Traffic Tycoon}?
	\item Why does \texttt{Map design} have so many decompositions into other tasks?
\end{itemize}

These are the type of the questions that we cannot answer just by looking at the GRL models. The model in Figure~\ref{fig:trafficsim} does not contain information about discussions that leads to the resulting elements of the model, such as various clarification steps for the naming, or alternatives that have been considered for the relationships. In this article we aim to address this shortcoming.

@@

Reasoning about which goals to pursue and actions to take is often referred to as \emph{practical reasoning}, and has been studied extensively in philosophy (e.g. \cite{Raz1978-RAZPR,walton1990}) and artificial intelligence \cite{bratman1987,atkinson2007}. One approach is to capture practical reasoning in terms of arguments schemes and critical questions~\cite{walton1990}. The idea is that an instantiation of such a scheme gives a presumptive argument in favor of, for example, taking an action. This argument can, then, be tested by posing critical questions about, for instance, whether the action is possible given the situation, and a negative answer to such a question leads to a counterargument to the original presumptive argument for the action.

A formal approach to persuasive and deliberative reasoning about goals and actions has been presented by Atkinson et al.~\cite{atkinson2007}, who define the \emph{practical reasoning argument scheme} (PRAS). PRAS follows the following basic argument structure.

\begin{itemize}
\item[] We have goal $G$,
\item[] Doing action $A$ will realize goal $G$,
\item[] Which will promote the value $V$
\item[] \textit{Therefore} 
\item[] We should perform action $A$
\end{itemize}

So, for example, we can say that 
\begin{itemize}
\item[] We have goal  \texttt{Generate traffic},
\item[]  \texttt{Keep same cars} will realize goal  \texttt{Generate traffic}, %TODO SG: where is this in the model?
\item[] Which will promote the value  \texttt{Simple design} %TODO SG: where is this in the model?
\item[] \textit{Therefore} 
\item[] We should perform action  \texttt{Keep same cars}
\end{itemize}

Practical reasoning is defeasible, in that conclusions which are at one point acceptable can later be rejected because of new information. Atkinson \emph{et al.}~\cite{atkinson2007} define a set of critical questions that point to typical ways in which a practical argument can be criticized by, for example, questioning the validity of the elements in the scheme or the connections between the elements. Some examples of critical questions are as follows.

\begin{enumerate}
\item Will the action bring about the desired goal?
%\item Does the goal promote the value stated?
\item Are there alternative ways of realizing the same goal?
\item Are there alternative ways of promoting the same value?
%\item Does doing the action have a side effect which demotes the value?
\item Does performing the action have a side effect which demotes some other value? %TODO SG: do you mean correlation link?
\item Does performing the action promote some other value?
%\item Does doing the action preclude some other action which would promote some other value?
\item Is the action possible?
\item Can the desired goal be realized?
\item Is the value indeed a legitimate value?
\end{enumerate}

These critical questions can point to new arguments that might counter the original argument. Take, for example, critical question 4: if we find that  \texttt{Keep same cars} actually negatively influences the value  \texttt{Realistic simulation}, we have a counterargument to the above argument.  %TODO SG: be careful with the concepts and semantics of links and elements in GRL. If you talk about side effect this indicates a correlation link. If you talk about impact or influence this indicates contribution link. In CQ4 you say side effect, here you say influence and in the model you show contribution link. Finally which one???? 
%TODO SG: Keep the same car in the model positively impacts the realistic simulation. But you say different here... VERY confusing.
Another way to counter an argument for an action is to suggest an alternative action that realizes the same goal (question 2) or an alternative goal that promotes the same value (question 3). For example, we can argue that  \texttt{Create new cars} also realizes the goal  \texttt{Generate traffic}, %TODO SG: where is this in the model?
which gives us a counterargument to the original argument -- to generate traffic by simply keeping the cars that disappear off the screen and have them wrap around to the other side of the screen -- that also follows PRAS. %TODO SG: what do you mean?

>>>>

Most of the examples in this article, as well as the topic of discussion in the transcripts we analyze, come from the traffic simulator design exercise. In this exercise, designers are provided a problem description, requirements, and a description of the desired outcomes. The problem description is given in full in Appendix~\ref{sect:designprompt}, and is summarized as follows: The client of the project is Professor E, who teaches civil engineering at UCI. It is the task of the designer to specify a system in which the professor can teach students how various theories around traffic lights works, such as queuing theory. To this end, a piece of software has to be developed in which students can create visual maps of an area, regulate traffic, and so forth. The original version of the problem descrption~\cite{UCIworkshop} is well known in the field of design reasoning since it has been used in a workshop\footnote{\url{http://www.ics.uci.edu/design-workshop/}}, and transcripts of this workshop have been analyzed in detail~\cite{Petre:2013:SDA:2535028}. Although the concepts of traffic lights, lanes, and intersections are common and appear to be simple, building a traffic simulator to represent these relationships and events in real time turns out to be challenging.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth]{img/example_small}
\caption{Partial GRL Model of Traffic Simulator Example (\todo{Marc}{Sepideh}{Turn into vector image, turn ``Dynamic simulation'' and ``Static Simulation'' into goals, turn ``Simulate'' into softgoal, add some resource}}
\label{fig:example-small}
\end{figure}

@@

Figure~\ref{fig:example-small} illustrates a partial GRL diagram from the traffic simulator design exercise (the full figure is shown in Figure~\ref{fig:transcripts:grl} and will be discussed in Section~\ref{sect:gmas}). An actor (\includegraphics[scale=1]{img/actor}) represents a stakeholder of a system or the system itself (\texttt{Traffic Tycoon}, Figure~\ref{fig:example-small}). Actors are holders of intentions; they are the active entities in the system or its environment who want goals to be achieved, tasks to be performed, resources to be available, and softgoals to be satisfied. Softgoals (\includegraphics[scale=1]{img/softgoal}) differentiate themselves from goals (\includegraphics[scale=1]{img/goal}) in that there is no clear, objective measure of satisfaction for a softgoal whereas a goal is quantifiable, often in a binary way. Softgoals (e.g. \texttt{Simulate}) are often related to non-functional requirements, whereas goals (such as  \texttt{Generate cars}) are related to functional requirements. Tasks (\includegraphics[scale=1]{img/task}) represent solutions to (or operationalizations of) goals and softgoals. In Figure~\ref{fig:example-small}, some of the tasks are  \texttt{Create new cars} and \texttt{Keep same cars}. In order to be achieved or completed, softgoals, goals, and tasks may require resources (\includegraphics[scale=1]{img/resource}) to be available (e.g., \todo{Marc}{Sepideh}{Add resource here}).

Different links connect the elements in a GRL model. AND, IOR (Inclusive OR), and XOR (eXclusive OR) decomposition links (\includegraphics[scale=1]{img/decomposition}) allow an element to be decomposed into sub-elements. In Figure~\ref{fig:example-small}, the goal \texttt{Generate cars} is XOR-decomposed to the tasks \texttt{Create new cars} and \texttt{Keep same cars}. Contribution links (\includegraphics[scale=1]{img/contribution}) indicate desired impacts of one element on another element. A contribution link has a qualitative contribution type or a quantitative contribution. Task  \texttt{Create new cars} has a \emph{help} qualitative contribution to the task \texttt{Dynamic}. Dependency links (\includegraphics[scale=1]{img/dependency}) model relationships between actors. \todo{Marc}{Sepideh}{Shall we add a dependency relation as well? They don't play a big role in our work..}

GRL is based on $i*$~\cite{Yu:1997:TMR:827255.827807} and the NFR Framework~\cite{chung2012non}, but it is not as restrictive as $i*$. Intentional elements and links can be more freely combined, the notion of agents is replaced with the more general notion of actors, i.e., stakeholders, and a task does not necessarily have to be an activity performed by an actor, but may also describe properties of a solution. GRL has a well-defined syntax and semantics. Furthermore, GRL provides support for providing a scalable and consistent representation of multiple views/diagrams of the same goal model (see~\cite[Ch.2]{Ghanavati2013} for more details). GRL is also linked to Use Case Maps via URNLink ((\includegraphics[scale=1]{img/urnlink}) which provides traceability between concepts and instances of the goal model and behavioral design models. Multiple views and traceability are a good fit with our current research: we aim to add traceability links between intentional elements and their underlying arguments.

@@

The GRL model in Figure~\ref{fig:example-small} shows the softgoals, goals, tasks and the relationship between the different intentional elements in the model. However, the rationales and arguments behind certain intentional elements are not shown in the GRL model. Some of the questions that might be interesting to know about are the following:

\begin{itemize}
	\item Why does actor \texttt{Traffic Tycoon} have only a single softgoal \texttt{Simulate}? Why is this, for instance, not connected to goal \texttt{Show Simulation}?
	\item What does \texttt{Keep same cars} mean?
	\item Why does task \texttt{Keep same cars} contribute positively to \texttt{Static simulation} and negatively to \texttt{Dynamic simulation}?
	\item Why does \texttt{Generate Cars} XOR-decompose into two tasks?
\end{itemize}

These are the type of the questions that we cannot answer just by looking at the GRL models. The model in Figure~\ref{fig:example-small} does not contain information about discussions that lead to the resulting model, such as various clarification steps for the naming, or alternatives that have been considered for the relationships. In this article we aim to address this shortcoming.

@@

Reasoning about which goals to pursue and actions to take is often referred to as \emph{practical reasoning}, and has been studied extensively in philosophy (e.g. \cite{Raz1978-RAZPR,walton1990}) and artificial intelligence \cite{bratman1987,atkinson2007}. One approach is to capture practical reasoning using arguments schemes and critical questions~\cite{walton1990}. Applying an argument schemes results in an argument in favor of, for example, taking an action. This argument can, then, be tested with critical questions about, for instance, whether the action is possible given the situation. A negative answer to such a question leads to a counterargument to the original argument for the action.

Atkinson et al.~\cite{atkinson2007} develop an argument scheme for practical reasoning, which they call the \emph{practical reasoning argument scheme} (PRAS). This argument scheme is as follows\footnote{The original argument schemes uses slightly different terminology: we have replaced \emph{value} with \emph{softgoal}. Our results do not depend on these adjustments, but they make the relation between PRAS and GRL more clear.}:

\begin{itemize}
\item[] We have goal $G$,
\item[] Doing action $A$ realizes to goal $G$,
\item[] Which will contribute positively to the softgoal $S$
\item[] \textit{Therefore} 
\item[] We should perform action $A$
\end{itemize}

The capital letters $G$, $A$, and $S$ are variables, which can be instantiated with concrete goals, actions, and softgoals. Once instantiated, we obtain an argument. We can for instance instantiate the argument scheme above with the elements of Figure~\ref{fig:example-small} as follows:
\begin{itemize}
\item[] We have goal \texttt{Static simulation},
\item[] Doing action \texttt{Keep same cars} realizes goal \texttt{Static simulation}, 
\item[] Which contribute positively to the softgoal \texttt{Simulate} 
\item[] \textit{Therefore} 
\item[] We should perform action \texttt{Keep same cars}.
\end{itemize}

Atkinson \emph{et al.}~\cite{atkinson2007} define a set of critical questions that point to typical ways in which a practical argument can be criticized by, for example, questioning the validity of the elements in the scheme or the connections between the elements. Some examples of critical questions are as follows.

\begin{enumerate}
\item Will the action realize the desired goal?
\item Are there alternative ways of realizing the same goal?
\item Are there alternative ways of contributing to the same softgoal?
\item Does performing the action contribute negatively to some other softgoal?
\item Is the action possible?
\item Can the desired goal be realized?
\item Is the softgoal indeed a legitimate softgoal?
\end{enumerate}

These critical questions can point to new arguments that might counter the original argument. Take, for example, critical question 1: if we find that  \texttt{Keep same cars} actually does not realize goal \texttt{Static Simulation}, we can form a counter-argument.

Another way to elaborate an argument for an action is to suggest an alternative action that realizes the same goal (question 2) or an alternative goal that contributes to the same softgoal (question 3). For example,  can argue that another goal \texttt{Semi-dynamic Simulation} also realizes the softgoal \texttt{Simulate}. \todo{Marc}{Sepideh}{Do you think our example is good? I think it is confusing. Perhaps we should just create an artificial example here, loosely based on the traffic simulator but where the model makes sense to the reader. Now the tasks are all a bit vague and I think nobody knows what they actually mean.}

### 2017-06-27 discussion.tex
many improvements (see email)

\section{Discussion}
\label{sect:goalmodeling:discussion}

@@

In this article, we develop the RationalGRL framework to trace back elements of GRL models to arguments using in discussions between stakeholders. The contributions of this article have become increasingly more formal. First, we analyze transcripts of meetings about the development of an information system. We count occurrences of argument schemes and critical questions, and categorize them by the effect of instantiating them. Then, we create a visual notation to link arguments and questions to elements of a goal model. If an argument for an element of a goal model is rejected, then the corresponding element is disabled. Similarly, if the argument is accepted, then the corresponding element is enabled. Finally, we formalize the argument schemes and critical questions in a logical framework. We propose a formal language to describe a GRL model, and we develop various algorithms for applying our argument schemes and critical questions.

Our framework is one of the first attempts that try to combine the formal theory of argumentation with the practical frameworks of goal modeling. History has shown that it is remarkably difficult to combine formal results with practice. By taking argument schemes as a starting point, we believe we have chosen the right level of formality, but as the open issues section shows, much work is left to be done.

>>>>

\section{Discussion}
\label{sect:discussion}

@@

The introduction of this article contains five requirements we identified for our framework. We use the conclusion to discuss how RationalGRL meets our initial requirements.

\paragraph{1. The argumentation techniques should be close to actual discussions stakeholders or designers have.}

We analyze a set of transcripts containing discussions about the architecture of an information system.

\paragraph{2. The framework must have the means to formally model parts of the discussion process.}
In order to generate goal models based on formalized discussions (requirement 2), we, first, formalize the list of arguments from requirement 1 in an argumentation framework. We formalize the critical questions as algorithms modifying the argumentation framework. We use argumentation techniques from AI in order to determine which arguments are accepted and which are rejected. We propose an algorithm to generate a GRL model based on the accepted arguments. This helps providing traceability links from GRL elements to the underlying arguments (requirement 3).

We implement our framework in an online tool called RationalGRL (requirement 4). The tool is implemented using Javascript. It contains  two parts, goal modeling and argumentation. The goal modeling part is a simplified version of GRL, leaving out features such as evaluation algorithms and key performance indicators. The argumentation part is new, and we develop a modeling language for the arguments and critical questions. The created GRL models in RationalGRL can be exported to jUCMNav~\cite{} %TODO SG: add citation 
 the Eclipse-based tool for GRL modeling, for further evaluation and analysis.

Our final contribution is a methodology on how to develop goal models that are linked to underlying discussions. The methodology consists of two parts, namely argumentation and goal modeling. In the argumentation part, one puts forward arguments and counter-arguments by applying critical questions. When switching to the goal modeling part, the accepted arguments are used to create a goal model. In the goal modeling part, one simply modifies goal models, which may have an effect on the underlying arguments. This might mean that the underlying arguments are no longer consistent with the goal models.

### 2017-06-27 examples.tex
many improvements (see email)

\section{Examples}
\label{sect:gmas:examples}

@@

\begin{figure}[ht!]
\centering
        \begin{tikzpicture}[->]
        \node (a0) [argNodeIN] at (-2,0) {
        	\argtext{AS3}{Actor \emph{System} has goal \emph{Simulate}}
        };
        \node (a1) [argNodeIN] at (-2,-1.5) {
        	\argtext{AS2}{Actor \emph{System} has task \emph{Dynamic simulation}}
        };
        \node (a2) [argNodeIN] at (-2,-3) {
        	\argtext{AS2}{Actor \emph{System} has task \emph{Static simulation}}
        };
        \node (a3) [argNodeOUT] at (-2,-5) {
        	\argtext{AS5}{Goal \emph{Simulate} AND-decomposes into \emph{Static simulation} and \emph{Dynamic simulation}}
        };
        \node (a4) [argNodeIN] at (-2,-7.5) {
        	\argtext{AS5}{Goal \emph{Simulate} OR-decomposes into \emph{Static simulation} and \emph{Dynamic simulation}}
        };
        \node[grl] (actor) at (2.3,-4) { 
        	\includegraphics[scale=0.5]{img/simulate_decomposition} 
        };
        \node[traceNodeGREEN] (trace1a) at (-.4,-.3) {};
        \node[traceNodeGREEN] (trace1b) at (2.5,-3.1) {};
        \node[traceNodeGREEN] (trace2a) at (-.4,-1.8) {};
        \node[traceNodeGREEN] (trace2b) at (3.5,-5.6) {};
        \node[traceNodeGREEN] (trace3a) at (-.4,-3.3) {};
        \node[traceNodeGREEN] (trace3b) at (1.3,-5.6) {};
        \node[traceNodeGREEN] (trace4a) at (-.4,-8.2) {};
        \node[traceNodeGREEN] (trace4b) at (2.5,-3.9) {};

\begin{pgfonlayer}{background}
         \path
    (a4) edge[attackLink] (a3);
\end{pgfonlayer}

	\path
	(a3) edge [CQLink, bend right=50] node [left,draw=none] {CQ10b} (a4)
    (trace1a) edge[traceLinkGREEN] (trace1b)
    (trace2a) edge[traceLinkGREEN] (trace2b)
    (trace3a) edge[traceLinkGREEN] (trace3b)
    (trace4a) edge[traceLinkGREEN, bend right] (trace4b);
\end{tikzpicture}
\caption{Argument schemes and critical questions (left), GRL model (right), and traceability link (dotted line) of the goal decomposition example.}
\label{fig:examples:decomposition}
\end{figure}

>>>>

\section{Examples}
\label{sect:examples}

@@

\begin{figure}[ht]
\begin{adjustwidth}{-3em}{}
        \begin{tikzpicture}[->]
        \node (a0) [argNodeIN] at (-2,0.5) {
        	\argtext{CQ3}{Task \emph{Dynamic simulation} is not relevant}
        };
        \node (a1) [argNodeIN] at (-2,-1.5) {
        	\argtext{AS2}{Actor \emph{System} has task \emph{Dynamic simulation}}
        };
        \node (a2) [argNodeIN] at (-2,-3) {
        	\argtext{AS2}{Actor \emph{System} has task \emph{Static simulation}}
        };
        \node (a3) [argNodeOUT] at (-2,-4.5) {
        	\argtext{AS5}{Goal \emph{Simulate} AND-decomposes into \emph{Static simulation} and \emph{Dynamic simulation}}
        };
        \node (a4) [argNodeIN] at (-2,-7.5) {
        	\argtext{AS5}{Goal \emph{Simulate} OR-decomposes into \emph{Static simulation} and \emph{Dynamic simulation}}
        };
        \node[grl] (actor) at (2.2,-4) { 
        	\includegraphics[scale=0.5]{img/simulate_decomposition} 
        };
        \node[traceNodeRED] (trace2a) at (-.4,-1.8) {};
        \node[traceNodeRED] (trace2b) at (3.2,-5.6) {};
        \node[traceNodeGREEN] (trace3a) at (-.4,-3.3) {};
        \node[traceNodeGREEN] (trace3b) at (1,-5.6) {};
        \node[traceNodeGREEN] (trace4a) at (-.4,-8.2) {};
        \node[traceNodeGREEN] (trace4b) at (2.2,-3.9) {};

\begin{pgfonlayer}{background}
         \path
    (a1) edge[attackLink] (a0)
    (a4) edge[attackLink] (a3);
\end{pgfonlayer}

	\path
	(a3) edge [CQLink, bend right=50] node [left,draw=none] {CQ10b} (a4)
	(a1) edge [CQLink, bend right=50] node [right,draw=none] {CQ3} (a0)
    (trace2a) edge[traceLinkRED] (trace2b)
    (trace3a) edge[traceLinkGREEN] (trace3b)
    (trace4a) edge[traceLinkGREEN, bend right] (trace4b);
\end{tikzpicture}
\end{adjustwidth}
\caption{Argument schemes and critical questions (left), GRL model (right), and traceability link (dotted line) of the example.}
\label{fig:examples:decomposition}
\end{figure}

### 2017-06-27 formalframework.tex
many improvements (see email)

\section{RationalGRL: the logical framework}
\label{sect:formalframework}

@@

Let us briefly review what we have done so far. Up until now we have achieve the following:
\begin{itemize}
\item We (informally) developed a set of argument schemes and critical questions in Section~\ref{sect:gmas}, by annotating transcripts.
\item We developed a visual notation to illustrate some argument schemes and critical questions in Section~\ref{sect:gmas:examples}.
\item We proposed a logical language consisting of atomic sentence to describe a GRL model in Section~\ref{sect:goalmodeling:logicallanguage}.
\item We formalized an ``argument'' as a set of atoms from our language, and we introduced \emph{argumentation semantics} to compute sets of accepted arguments in Section~\ref{sect:goalmodeling:argumentationsemantics}.
\end{itemize}

We see that we are becoming increasingly more formal in each step of the progress. What we haven't formalized properly yet are the argument schemes and critical questions, which is what we turn to now.

Both the application of an argument schemes and the answering of critical questions are \emph{procedural} or \emph{dialectic} in nature. Therefore, a natural choice of formalizing them seems to develop procedures, or algorithms for them, which is what we do here.

In the following algorithms, we assume the following global variables:

>>>>

\section{RationalGRL: Logical Framework}
\label{sect:formalframework}

@@

Let us briefly take stock:
\begin{itemize}
\item We developed an initial set of argument schemes and critical questions in Section~\ref{sect:gmas}, by annotating transcripts.
\item We developed a visual notation to illustrate some argument schemes and critical questions in Section~\ref{sect:examples}.
\item We proposed a logical language consisting of atomic sentence to describe a GRL model in Section~\ref{sect:goalmodeling:logicallanguage}.
\item We formalized an ``argument'' as a set of atoms from our language, and we introduced \emph{argumentation semantics} to compute sets of accepted arguments in Section~\ref{sect:goalmodeling:argumentationsemantics}.
\end{itemize}

In this subsection we develop algorithms for applying argument schemes and critical questions. In the following algorithms, we assume the following global variables:

### 2017-06-27 gmas.tex
many improvements (see email)

In this section, we develop a set of \emph{argument schemes for goal modeling} and associated critical questions. We start from an initial list that we derive from PRAS, containing argument schemes and critical questions that specific to elements and relationships of GRL. We then refine this list incrementally by annotating transcripts, meaning we identify and count instantiations of arguments schemes. If we find new argument schemes or critical questions, we add them to the list. If some of them do not occur at all, we remove them from the list. We then manually construct goal models from the arguments and counter arguments we find in the transcripts, and add traceability links between the goal model and underlying arguments. An example is shown in Figure~\ref{fig:transcripts:grl}, which is the same as the example in Figure~\ref{fig:trafficsim}, but now including tracability links to underlying arguments. A red dot indicates the underlying argument is rejected, and a green dot means the underlying argument is accepted.

The final list of argument schemes and critical questions that we end up with after our analysis is shown in Table~\ref{table:argument-schemes}. The initial list of argument schemes, i.e., those we start with before annotating the transcripts, consists of AS1-AS4, AS6-AS9 (Table~\ref{table:argument-schemes}), and their corresponding critical questions. The first four argument schemes (AS0-AS4) are arguments for an element of a goal model, the next seven (AS5-AS11) are about relationships, the next two (AS12-AS13) are about intentional elements in general, and the last is (Att) is a generic counterargument for any type of argument that has been put forward.

We found that answering critical questions can have varying effects on the model, and for each critical question, the right column in Table~\ref{table:argument-schemes} shows the effect of answering the critical questions affirmatively. Answering a critical questions can create an argument disabling the corresponding GRL element of the attacked argument scheme (\textsf{DISABLE}); it can create an argument introducing a new GRL element (\textsf{INTRO}); it can replace the GRL element corresponding to the original argument (\textsf{REPLACE}), or it can simply attack an argument directly (\textsf{ATTACK}).

@@

\paragraph{Experimental Setup} The assignment used for the experiments is to design a traffic simulator. Participants were asked to use a think-aloud method during the design session. The assignment was slightly adjusted to include several viewpoints as end products in order to conform to the course material~\cite{Bass:2012:SAP:2392670}. The full problem descriptions can be found in Appendix~\ref{ch:designprompt}. All groups were instructed to apply the \emph{functional architecture method}, focusing on developing the \emph{context}, the \emph{functional}, and the \emph{informational} viewpoints of the traffic simulator software. The students had two hours for the tasks, and the transcripts document the entire discussion. The details of the transcripts are shown in Table~\ref{table:transcripts:info}.

@@

All original transcripts, annotations, and models are available on the following website:
\begin{quote} \url{http://www.github.com/marcvanzee/RationalArchitecture}
\end{quote}
in the folder ``Ch4 transcript analysis''. We also provide excerpts of the annotation in Appendix~\ref{sect:transcripts:excerpts}, and most of the examples we use in this article come from the transcripts.

>>>>

In the previous section we provided an overview of RationalGRL. In the example of the methodology, we discussed various argument schemes and critical questions. In this section, we discuss our argument schemes and critical questions in more detail. It is important to note that the list of argument schemes we obtain in this section is not exhaustive. It is an initial list that we obtained by annotating transcripts, but our framework is fully extensible, meaning that new argument schemes and critical questions can be added depending on the problem domain.

The list of argument schemes and critical questions that we obtained from our analysis is shown in Table~\ref{table:argument-schemes}. The first four argument schemes (AS0-AS4) are arguments for an element of a goal model, the next seven (AS5-AS11) are about relationships, the next two (AS12-AS13) are about intentional elements in general, and the last is (Att) is a generic counterargument for any type of argument that has been put forward.

As we already discussed in the methodology, we found that answering critical questions can have varying effects on the model, and for each critical question, the right column in Table~\ref{table:argument-schemes} shows the effect of answering the critical questions affirmatively. Answering a critical questions can create an argument disabling the corresponding GRL element of the attacked argument scheme (\textsf{DISABLE}); it can create an argument introducing a new GRL element (\textsf{INTRO}); it can replace the GRL element corresponding to the original argument (\textsf{REPLACE}), or it can simply attack an argument directly (\textsf{ATTACK}).

@@

\paragraph{Experimental Setup} The assignment used for the experiments is to design a traffic simulator. Participants were asked to use a think-aloud method during the design session. The assignment was slightly adjusted to include several viewpoints as end products in order to conform to the course material~\cite{Bass:2012:SAP:2392670}. The full problem descriptions can be found in Appendix~\ref{sect:designprompt}. All groups were instructed to apply the \emph{functional architecture method}, focusing on developing the \emph{context}, the \emph{functional}, and the \emph{informational} viewpoints of the traffic simulator software. The students had two hours for the tasks, and the transcripts document the entire discussion. The details of the transcripts are shown in Table~\ref{table:transcripts:info}.

@@

All original transcripts, annotations, and models are available on the following website \todo{Marc}{Marc}{Change URL and put everything together with the tool}:
\begin{quote} \url{http://www.github.com/marcvanzee/RationalArchitecture}
\end{quote}
We also provide excerpts of the annotation in Appendix~\ref{sect:transcripts:excerpts}, and most of the examples we use in this article come from the transcripts.
